---
title: "CV"
description: |
  Work Experience 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

```{r}
ggplot(data = diamonds, aes(x = carat, y = price)) +
  geom_point(aes(color = price))
```





### Education

Bren School | UCSB | Santa Barbara, California 
M.S., Environmental Data Science, Expected Graduation June 2022

Boston University | Boston, Massachusetts
B.A., Marine Sciences, May 2021

Santa Barbara Senior High School | Santa Barbara, California
High School Diploma, June 2017



### Publications

Rotjan, R.D., N.E. Ray, B.R.C. Kennedy, I. Cole, I.F. Trumble, G. McCarthy, K. Castro, A. Bhardwaj, M. Edens, I. Karageorge, C. Klawon, H. Kruh-Needleman, R. Perez, C. Roberts, A. Volk, J. Torres, and J. Morey. Indirect indications of dietary shifts following climate-induced disturbance on a coral reef. In preparation for submission to the Marine Ecology Progress Series. 

Cole, I.*, E. Martin*, J. Jaskiel, C. Hernandez, J. Witting, J. Llopiz, and R.D. Rotjan. Studying Global Change in a Natural Climate Laboratory: Are Larval Fishes Exhibiting a Distribution Shift in Response to ENSO Events? *= co-first authors. In preparation for submission to Science of the Total Environment. 


### Research Experience

Research Assistant | Work-Study | Rotjan Marine Ecology Lab | Boston University                          09 2018 – Present
http://sites.bu.edu/rotjanlab/members/ 
20+ hours a week
  -Examine deep sea, open ocean, and coral reef fish larvae in the Phoenix Islands Protected Area as part of a 10-year project determining how community assemblages shift with ENSO cycles (~45 fish taxa and 28 invertebrate taxa)
  -Maintain temperate and tropical aquarium systems. Including husbandry of corals, anemones, and mantis shrimp, water quality testing/monitoring, feeding, and cleaning
  -Mentor fellow undergraduates in husbandry and lab protocols as they began doing research, and how use the microscopes, sort zooplankton, and identify larval fish 
  -Present my work at Rotjan Lab meetings, help to edit and write manuscripts for submission to peer-reviewed journals


