---
title: "Tidyverse Cleaning"
description: |
 A few helpful functions from the Tidyverse 
author:
  - name: Allie Cole
    url: {}
date: 08-16-2021
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(palmerpenguins)
```

## Practice some tidyverse functions for data wrangling using `palmerpenguins` as an example

- `filter()`
- `select()`
- `mutate()`
- `case_when()`
- `group_by() %>% summarize()` 
- piping in sequence


### dplyer::filter() to keep ot exclue rows based on conditions 

1. Exact match (single condition)
 - keep only observations where the island is Biscoe 

```{r}
ex_1 <- filter(.data = penguins,
               island == "Biscoe")
unique(ex_1$island) #to check and make sure it worked
```


2. `and` statement within filter()
- only keep observations if the penguins are chinstrap and the year is 2008 

```{r}
ex_2 <- penguins %>% 
  filter(species == "Chinstrap",
           year == 2008) # coudl also use & instead of , or filter each seperate
```


3. Exclude Gentoos 
- keep observations if teh penguin species are NOT gentoo

```{r}
ex_3 <- penguins %>% 
  filter(species != "Gentoo")
```


Practice:
- Create a subset from penguins that only caintains adlieie with a flipper length greater than or equal to 185mm 

```{r}
prac_1 <- penguins %>%  
  filter(species == "Adelie", flipper_length_mm >= 185)
```


4. Using teh %in% operator
- to look for any matches within a vector, and return true if teh value does exist for our row 

we want to keep observations where the island is dream or torgerson 

```{r}
ex_4 <- penguins %>% 
  filter(island %in% c("Dream", "Torgersen")) # add a ! before island to get anything not in this vector 
```

### Select() to select or exclude columns

5. Keep a single column by name 

- Only want to keep teh species column 

```{r}
ex_5 <- penguins %>% 
  select(species) %>% 
  pull(species) #now its all the values in a vector not a df 

#reminder... rewrite without teh pipe 

ex_5 <- select(.data = penguins, species)
```

6. Keep multiple columns by name 

- Create a subset that contains species, bill length, through flipper length and year 

```{r}
ex_6 <- penguins %>% 
  select(species, bill_length_mm:flipper_length_mm, year)

```


7. Chose or exclude columns based on position 

```{r}
ex_7 <- penguins %>%  
  select(-4)
```

8. A piped sequence 
- Write a piped sequence where you first only keep columns for species body mass and island and second filter to only include observations if body mass if greater than 3600

```{r}
ex_8 <- penguins %>% 
  select(species, body_mass_g, island) %>% 
  filter(body_mass_g >= 3600)
```

### mutate() add or update existing columns while keeping the ones already there 

9. Add a column that has body mass converted to kg 

```{r}
ex_9 <- penguins %>% 
  mutate(body_mass_kg = body_mass_g / 1000) #can add more statements on if you want 
```


10. mutate() to update an existing column 

```{r}
ex_10 <- penguins %>% 
  mutate(island_hooray = str_replace(island, pattern = "sen", replacement = "HOORAY!!")) #o this is great for fixing spelling mistakes 
```

11. Update the class of the bill_depth_mm column to character 

```{r}
ex_11 <- penguins %>% 
  mutate(bill_depth_mm = as.character(bill_depth_mm))

class(ex_11$bill_depth_mm) #testing
```


### case_when(): your favorite if-else 

12. Create a new column that contains small if penguin flipper length is less than 185, medium if otherwise 

```{r}
ex_12 <- penguins %>% 
  mutate(binned_size = case_when(
    flipper_length_mm < 185 ~ "small", 
    flipper_length_mm >= 185 ~ "medium"
  ))
```


Practice: starting fro penguins create a new column called bird_bill_bin that contains 
- bill length is less than 38mm then "short bill"
- if >= to 38mm & < 43mm "meduim bill"
- if >= 43 then is "long bill

```{r}
prac_2 <- penguins %>% 
  mutate(bird_bill_bin = case_when(
    bill_length_mm < 38 ~ "small bill", 
    bill_length_mm >= 38 & bill_length_mm < 43 ~ "medium bill", 
    bill_length_mm > 43 ~ "long bill"
  ))
```


### group_by () %>% summarize() to find summary values by group and retunr tehm in a nice summary table 

13. Find the mean of the bill_depth_mm column by species 

```{r}
mean(penguins$bill_depth_mm, na.rm = TRUE)

ex_13 <- penguins %>%  
  group_by(species, island) %>% #can add as many of the grouping variables as you want
  summarize(mean_flipper_length = mean(bill_depth_mm, na.rm = TRUE), 
            sd_flipper_length = sd(bill_depth_mm, na.rm = TRUE))

```

Practice 

In a single pipe sequence 

- Keep penguins with a body mass greater than 33500 grams 
- Only keep columns for species, flipper length and sex
- Create a summary table that contains teh mean , sd, max and min value of flipper length by species and sex 

```{r}
prac_3 <- penguins %>% 
  filter(body_mass_g > 3500) %>%
  select(species, flipper_length_mm, sex) %>% 
  group_by(species, sex) %>% 
  summarize(mean_flipper_length = mean(flipper_length_mm, na.rm = TRUE), 
            sd_flipper_length = sd(flipper_length_mm, na.rm = TRUE), 
            max_flipper_length = max(flipper_length_mm, na.rm = TRUE), 
            min_flipper_length = min(flipper_length_mm, na.rm = TRUE))
```









